\section{Conclusion}
\label{sec:conclusion}

Agent memory systems inherit a convenient assumption: that a model can say what is
worth remembering. Under controlled content provenance the assumption breaks in a structured
way: self-rated importance is excellent for explicit content and, on the
construct-valid benchmarks, at or below chance for silently inferred content
at every scale and post-training stage we measured, in three model families,
with the cue-density boundary mapped in App.~\ref{app:replications}. A J-lens-derived
workspace readout recovers much of the signal that self-report loses on evoked
content and causally feeds memory-based answers; used as a write gate (with
PRG, a provenance-routed extension, for tight budgets on mixed streams), it
matches the canonical 1--10 rating at zero probe cost while leading on the
construct-valid slice, selects a memory budget at up to selection-oracle
quality, and beats the binary importance probe on a mixed-provenance
stream. The failure is remediable: post-training never connects the verbal channel to
the workspace because none of its objectives depend on that connection, and
roughly 500 self-supervised steps suffice to establish it. Beyond agents, we read this as a post-training lesson: report-channel objectives
do not by themselves calibrate self-report against internal state, but
self-distillation from internal signals cheaply can.

\textbf{Limitations.}
Our benchmarks are synthetic and single-word-concept by design; control over
provenance is the point, and it is not available elsewhere: on the two
standard naturalistic benchmarks (LoCoMo, LongMemEval) every candidate is
surface content by construction, no question-blind channel separates from
random there, and the evidence oracle doubles all of them
(App.~\ref{app:locomo}). A naturalistic stream rich in silent inference does
not yet exist and is the natural next instrument; embedding our episodes in
real dialogue carriers preserves the dissociation at 7B
(App.~\ref{app:locomo}), so the obstacle is benchmark construction, not
transfer of the effect. Each benchmark is written by one LLM generator, so we replicated across
three: the workspace readout is stable across the generators that produce the
construct-valid design ($0.53$--$0.66$; on gpt-4o's weakly evoked bridges both
channels attenuate), while the verbal channel's calibration on silent
inferences swings in both directions, anti-calibrated on sparse-cue episodes
at every scale and recovering on densely cued ones at 7B
(App.~\ref{app:replications}); the anti-calibration claim is scoped
accordingly, and the workspace channel's stability is precisely its case.
Episode counts (52--120 per benchmark) give adequate power only for the
pre-specified contrasts. The Leverage readout's compositional-regime win holds
for Qwen but not OLMo-2-7B; the 3B downstream gap on Decoupled-class tasks is bounded by
recall-time composition, not selection; and in vision--language probes, visually presented
identity behaves like the explicit regime (near-perfect ask-time verbal access
in both Qwen2-VL-2B and LLaVA-1.5-7B), while its word-level workspace trace is
family-dependent, absent at 2B and partial in LLaVA-7B
(App.~\ref{app:multimodal}), so our claims are about language-derived
inference. Post-training claims are scoped to our stage sweep and to utility-salience. Our
results establish the value of training-free internal readouts for write
admission; whether supervised hidden-state probes do better is open (they would
require labeled training data that the write-gate setting does not naturally
provide). Full deviations from the pre-specified
predictions are listed in App.~\ref{app:prereg}.

\textbf{Outlook.}
Three extensions look immediate: widening PRG's routing to the full
availability family (the gradient-based Leverage readout for compositional
content), metacognitive alignment at scale and for other self-knowledge
channels, and evaluating workspace gating inside production memory stacks on
naturalistic long-horizon streams.
