\section{Benchmarks and Measurement Protocol}
\label{sec:setup}

\subsection{Problem Setting}
\label{sec:problem}

Let $\mathcal{M}$ be a frozen language model and let an \emph{episode} be a
tuple $e=(x, C, q, a)$: a context passage $x$, a candidate set
$C=\{c_1,\dots,c_m\}$ of concepts that could be written to memory, and a probe
question $q$ with gold answer $a$ that arrives only after $x$ has left the
context window. A \emph{write-admission policy} with budget $k$ is a map
\begin{equation}
\pi_k : (x, C) \mapsto K \subseteq C, \qquad |K| \le k,
\label{eq:policy}
\end{equation}
constrained to observe only the encoding-time state, with access to neither
$q$ nor $a$.
At recall, the model answers from the kept set alone, and the policy's value is
its downstream accuracy,
\begin{equation}
\mathrm{Acc}_k(\pi) \;=\; \mathbb{E}_{e}\!\left[\,
\mathbf{1}\{\mathcal{M}(q_e \mid \pi_k(x_e, C_e)) \simeq a_e\}\,\right],
\label{eq:acc}
\end{equation}
where $\simeq$ denotes answer match under the grader of App.~\ref{app:repro}.
Each candidate carries a latent utility label $y(c)\in\{0,1\}$ ($y{=}1$ iff $c$
is required to answer $q$; we call such items \emph{load-bearing}). A policy
reduces to a scoring function $S(c; x)\in\mathbb{R}$ ranked within episodes,
and its selection quality is the probability of ranking a load-bearing item
above a non-load-bearing one,
\begin{equation}
\mathrm{AUC}(S) \;=\; \Pr\big[\,S(c^{+}; x^{+}) > S(c^{-}; x^{-})\,\big]
\;+\; \tfrac{1}{2}\,\Pr\big[\,S(c^{+}; x^{+}) = S(c^{-}; x^{-})\,\big],
\label{eq:auc}
\end{equation}
where $(c^{+}, x^{+})$ and $(c^{-}, x^{-})$ are drawn uniformly from the pooled
item populations with $y{=}1$ and $y{=}0$ respectively, and exact ties (common
for the rank-valued scores below) receive half credit.

\textbf{Provenance.} The variable this paper manipulates is the
\emph{provenance} of the load-bearing concept,
\begin{equation}
\rho(c, x) \;=\; \mathbf{1}\{\text{$c$ occurs as a surface form in } x\},
\label{eq:provenance}
\end{equation}
distinguishing content that is \emph{stated} ($\rho{=}1$) from content that is
\emph{silently inferred} ($\rho{=}0$): derivable from $x$ but never written.
Our central question is how the reliability of different scoring channels
depends on $\rho$.

\subsection{Benchmarks}
\label{sec:benchmarks}

\textbf{Why controlled construction is necessary.}
Answering the question above imposes two requirements on a benchmark: the
provenance of the load-bearing item must vary across episodes, so that
episodes exist whose outcome hinges on a $\rho{=}0$ concept, and which
candidate is load-bearing must be known independently of every channel under
test. Existing naturalistic memory benchmarks satisfy neither, and not by
accident but by construction: their candidate memories are surface content.
In LongMemEval \citep{wu2025longmemeval} the natural memory units are the
conversation turns themselves, verbatim substrings of the context, so every
candidate has $\rho{=}1$; in LoCoMo \citep{maharana2024locomo} the annotated
observations summarize stated dialogue content (a median of $83\%$ of a
note's content words appear verbatim in the source session). On such streams
the regime in which the channels could dissociate never occurs, and
empirically no question-blind admission policy separates from random
selection on either benchmark, while an oracle given the evidence location
doubles their accuracy (App.~\ref{app:locomo}). Measuring the effect of
$\rho$ therefore requires constructing the variation deliberately, which is
what our benchmarks do.

Each episode contains one load-bearing concept, two to three vivid but useless
\textsc{distractor} concepts present in $x$, and neutral \textsc{filler} items.
The four benchmarks differ only in the provenance and role of the load-bearing
item:
\textbf{Explicit} ($\rho{=}1$; stated in text; $88$ episodes / $417$ items);
\textbf{Evoked} ($\rho{=}0$; a one-hop silent bridge, e.g.\ $x$ mentions Oslo
and $c^{+}$ is \emph{norway}; $75/352$);
\textbf{Decoupled} ($\rho{=}0$; as Evoked, but the gold answer $a$ is one
further knowledge hop from $c^{+}$, with $a \notin C$, $a \notin x$, and $q$
sharing no content word with $x$ except a person's name, so neither the probe
nor any stored surface item reveals $a$; $68/335$, plus a $120$-episode scaled
version, \textbf{Decoupled-L}); and
\textbf{Compositional} ($\rho{=}0$; the bridge is only weakly evoked and
matters solely through a two-hop composition; $52/261$; its probes carry
surface anchors, so it is construct-valid for encoding-time prediction only,
App.~\ref{app:benchmarks}).
We call Decoupled and Compositional \emph{construct-valid}: by construction,
neither the probe nor any stored surface item reveals the answer, so recall
success must pass through the inferred bridge. Evoked is
\emph{answer-coupled}: its bridge is also the gold answer, which makes it the
most sensitive design but lets stored surface content occasionally substitute
for the bridge.
Benchmarks are LLM-generated under these constraints and mechanically
re-validated (bridge absent from $x$; Decoupled: $a$ absent from $x$ and $C$;
probe anchor-free), then concept-deduplicated; construction details and a
leakage audit appear in App.~\ref{app:benchmarks}. As a sanity check, the
surface predictor $S{=}\rho$ carries no signal on Explicit
($\mathrm{AUC}{=}0.48$), and dropping every audit-flagged near-miss episode
changes no conclusion. On the inferred-regime benchmarks the planted bridge
is by design nearly the only context-absent candidate, so bare absence
($1{-}\rho$) trivially separates it at the episode level; the workspace
signal is not bare absence: among purely absent candidates the readout ranks
the evoked bridge far above matched absent foils, and the confound analyses
of App.~\ref{app:vrobust} separate evocation from absence and from semantic
relatedness.

\subsection{Two Channels for Scoring Utility}
\label{sec:channels}

\textbf{Workspace channel.}
The J-lens construction \citep{anthropic2026workspace} established that
verbalizable intermediate concepts occupy a decodable residual-stream
workspace: reading intermediate layers through the model's own unembedding
surfaces concepts the model is carrying, and swapping the corresponding
directions swaps downstream behavior. We inherit exactly that measurement
and contribute its use as a write-admission signal. Let
$h_\ell(x)\in\mathbb{R}^{d}$ be the residual-stream state at the final position
of $x$ in layer $\ell$, and let
$z_\ell(x) = W_U\,\mathrm{LN}\!\left(h_\ell(x)\right) \in
\mathbb{R}^{|\mathcal{V}|}$ be its logit-lens decoding through the model's own
final norm and unembedding \citep{nostalgebraist2020logitlens}. Writing
$\mathrm{rank}_\ell(t)$ for the rank of token $t$ in $z_\ell(x)$, the workspace
salience of a concept $c$ is its peak reciprocal rank,
\begin{equation}
W(c; x) \;=\; \max_{\ell \le L} \;\max_{t \in T(c)}\;
\frac{1}{\mathrm{rank}_\ell(t)},
\label{eq:wrr}
\end{equation}
where $T(c)$ is the small set of case-variant first tokens of $c$
(App.~\ref{app:repro}). $W$ is training-free, requires no forward passes beyond
encoding, and is read blind to $q$; reciprocal rank rather than probability
keeps the score scale-free across a ${\sim}150$k-token vocabulary. The double
maximum over layers and case-variant first tokens is an empirical aggregation
choice, not a derived optimum: it is motivated by scale-invariance, and we
compare $W$ against a conditional log-likelihood score and a trained
future-token lens (\S\ref{sec:extensions}, App.~\ref{app:vrobust}) rather than
claim optimality for the aggregation itself.

\textbf{Verbal channel.}
The same model is asked directly. With yes/no token sets $Y, N$ and the
importance probe $u(c, x)$ of App.~\ref{app:vrobust},
\begin{equation}
V(c; x) \;=\;
\frac{\sum_{t\in Y} p_{\mathcal{M}}\!\left(t \mid u(c,x)\right)}
     {\sum_{t\in Y\cup N} p_{\mathcal{M}}\!\left(t \mid u(c,x)\right)},
\label{eq:v}
\end{equation}
a binary-probability proxy for the Generative-Agents-style importance score;
the canonical 1--10 rating variant is evaluated in App.~\ref{app:vrobust} and
is likewise at or below chance on the construct-valid benchmarks. Base
checkpoints are additionally probed with a template-free variant $V_{raw}$.
Baselines include embedding relevance $E(c;x)=\cos(\phi(c),\phi(x))$ under a
sentence encoder $\phi$ (bge-small-en-v1.5, CLS pooling), recency, frequency, a surface-feature fusion proxy
(App.~\ref{app:fusion}), and three paraphrases of the probe $u$.

\subsection{Evaluation Protocol}
\label{sec:protocol}

The primary predictive metric is Eq.~\eqref{eq:auc} per channel, benchmark, and
model, with item-level bootstrap CIs in tables and episode-cluster bootstrap
CIs ($B{=}4{,}000$, resampling episodes) for all headline contrasts; the four
pre-specified primary contrasts ($W{-}V$ on Decoupled at each size) are
Bonferroni-corrected. By loose analogy with metacognitive-sensitivity analysis
\citep{fleming2014how}, each cell is summarized by a descriptive ratio
\begin{equation}
\Ms \;=\; \frac{\mathrm{AUC}(V) - 0.5}{\mathrm{AUC}(W) - 0.5},
\label{eq:ms}
\end{equation}
reported only where $\mathrm{AUC}(W)\ge 0.55$; $\Ms<0$ indicates that the
verbal report ranks truly load-bearing items below the rest even though the
signal is present in the system ($\Ms$ is negative throughout the
construct-valid regime; on answer-coupled Evoked it can be positive). We read
$\Ms$ qualitatively only: the denominator $\mathrm{AUC}(W){-}0.5$ is small near
chance, so the ratio is unstable there (hence the $0.55$ floor) and we
interpret its sign and coarse magnitude, not its precise value. Throughout, ``below chance'' refers to
ranking AUC, not probabilistic calibration in the forecasting sense.
Downstream, Eq.~\eqref{eq:acc} is estimated per policy and budget with
per-episode correctness logged for every condition; policies are compared on
identical episodes with exact McNemar tests, bounded by no-memory and
full-context references. Models: Qwen2.5 0.5B--7B \citep{qwen2025}, OLMo-2
1B/7B with released SFT/DPO/RLVR stages \citep{olmo2}, Mistral-7B
\citep{jiang2023mistral}; greedy decoding; a single fixed harness for all
checkpoints so that lens bias cancels in contrasts.
